Parallel Adaptive Importance Sampling

### 2016-06-14 PAIS.tex
fixed some typos in the introduction+preliminaries

Many advances have been made in the field of MCMC to design ever more
complex methods that propose moves more intelligently, leading to
rapidly converging methods. Function space versions of standard methods
such as the random walk Metropolis-Hastings (RWMH) algorithm or the
Metropolis adjusted Langevin algorithm (MALA), whose convergence rates
are independent of dimension have been
developed\cite{cotter2013mcmc}. The hybrid (or Hamiltonian) Monte
Carlo (HMC) method uses Hamiltonian dynamics in order to propose and
accept moves to states which are a long way away from the current
position\cite{sexton1992hamiltonian}, and function space analogues of
this have also been proposed\cite{beskos2011hybrid}. Riemann
manifold Monte Carlo methods exploit the Riemann geometry of the
parameter space, and are able to take advantage of the local structure
of the target density to produce more efficient MCMC
proposals\cite{girolami2011riemann}. This methodology has been
successfully applied to MALA-type proposals and methods which exploit
even higher order gradient information\cite{bui2014solving}.  These
methods allow us explore the posterior distribution more fully 
with fewer iterations.

@@

As such, the development of efficient parallel MCMC algorithms is an
important area for research. Since MCMC methods can be trivially
parallelised by simply running many independent chains in parallel,
the focus needs to be on the development of methods which gain some
added benefit through parallelism. One class of parallel MCMC method
uses multiple proposals, with only one of these proposals being
accepted. Examples of this approach include multiple try
MCMC\cite{liu2000multiple} and ensemble MCMC\cite{neal2011mcmc}. In
\cite{calderhead2014general}, a general construction for the
parallelisation of MCMC methods was presented, which demonstrated
speed ups of up to two orders of magnitude when compared with serial methods.

In this paper, we present a framework for parallelisation of
importance sampling, which can be built around many of the current
Metropolis-based methodologies in order to create an efficient target
proposal from the current state of all of the chains in the
ensemble. The idea is to consider the parallel chains as an ensemble,
and to resample using a transformation based on optimal transport.
Samplers based on optimal transport have also been considered in
\cite{el2012bayesian}.

In Section \ref{Sec:Prelim} we introduce some mathematical
preliminaries upon which we will later rely. In Section \ref{Sec:PAIS}
we present the general framework of the PAIS algorithm. In Section
\ref{Sec:adapt} we consider adaptive versions of PAIS which
automatically tune algorithmic parameters concerned with the proposal
distributions. In Section \ref{sec:AMR} we introduce the approximate
multinomial resampling (AMR) algorithm which is a less accurate but
faster alternative to the ETPF resampler. In Section \ref{Sec:Num} we
present some numerical examples, before a brief conclusion and
discussion in Section \ref{Sec:Conc}.

@@

The parameter $u$ is observed
through the observation operator $\mathcal{G}:X \to\mathbb{R}^d$. 
Since observations are never perfect, we
assume that these measurements $\data$ are subject to Gaussian noise, so that
\begin{equation}\label{eqn:obs}
	\data = \mathcal{G}(u) + \varepsilon, \qquad \varepsilon \sim \mu_{\varepsilon} = \mathcal{N}(0,\Sigma).
\end{equation}
For example, if $u$ are the rates of reactions in a chemical system, $\G$ might return the times at which each reaction occurs, or some summary of this information.

@@

The Ensemble Transform (ET) method proposed by Reich \cite{reich2013nonparametric} makes use of optimal transportation as described in \cite{villani2003topics,villani2008optimal}. The transform takes a sample of weighted particles $\{y_i\}_{i=1}^M$ from $\mu_Y$ and converts it into a sample of evenly weighted particles $\{x_i\}_{i=1}^M$ from $\mu_X$, by means of defining a coupling $T^*$ between $Y$ and $X$. Given that a trivial coupling $T^t$ always exists in the space of transference plans, $\Pi(\mu_{X}, \mu_{Y})$, we can find a coupling $T^*$ which maximises the correlation between $X$ and $Y$ \cite{cotter2012ensemble}. This coupling is the solution to a linear programming problem in $M^2$ variables with $2M-1$ constraints. Maximising the correlation ensures that the new sample is as much like the original sample as possible with the additional property that the sample is evenly weighted.

@@

\subsection{Deficiencies of Metropolis-type MCMC schemes}
All MCMC methods are trivially parallelisable. One can take a method
and simply implement it simultaneously over a set of processors in an ensemble. All
of the states of all of the ensemble member can be recorded, and in the time
that it takes one MCMC chain to draw $N$ samples, $M$ ensemble members
can draw
$NM$ samples.

However, we argue that this is not an optimal
scenario. First of all, unless we have a lot of information about the
posterior, we will intialise the algorithm's initial state in the
tails of the distribution. The samples that are initially made as the
algorithm finds its way to the mode(s) of the distribution cannot be
considered to samples from the target distribution, and must be thrown
away. This process is known as the burn-in. In a trivially
parallelised scenario, each ensemble member must perform this process
independently, and therefore mass parallelisation makes no inroads to
cutting this cost.

>>>>

Many advances have been made in the field of MCMC to design ever more
complex methods that propose moves more intelligently, leading to
rapidly converging methods. Function space versions of standard methods
such as the random walk Metropolis-Hastings (RWMH) algorithm or the
Metropolis adjusted Langevin algorithm (MALA), whose convergence rates
are independent of dimension have been
developed\cite{cotter2013mcmc}. The hybrid (or Hamiltonian) Monte
Carlo (HMC) method uses Hamiltonian dynamics in order to propose and
accept moves to states which are a long way away from the current
position\cite{sexton1992hamiltonian}, and function space analogues of
this have also been proposed\cite{beskos2011hybrid}. Riemann
manifold Monte Carlo methods exploit the Riemann geometry of the
parameter space, and are able to take advantage of the local structure
of the target density to produce more efficient MCMC
proposals\cite{girolami2011riemann}. This methodology has been
successfully applied to MALA-type proposals and methods which exploit
even higher order gradient information\cite{bui2014solving}.  These
methods allow us to explore the posterior distribution more fully 
with fewer iterations.

@@

As such, the development of efficient parallel MCMC algorithms is an
important area for research. Since MCMC methods can be naively
parallelised by simply running many independent chains in parallel,
the focus needs to be on the development of methods which gain some
added benefit through parallelism. One class of parallel MCMC method
uses multiple proposals, with only one of these proposals being
accepted. Examples of this approach include multiple try
MCMC\cite{liu2000multiple} and ensemble MCMC\cite{neal2011mcmc}. In
\cite{calderhead2014general}, a general construction for the
parallelisation of MCMC methods was presented, which demonstrated
speed ups of up to two orders of magnitude when compared with serial methods.

In this paper, we present a framework for parallelisation of
importance sampling, which can be built around many of the current
Metropolis-based methodologies in order to create an efficient target
proposal from the current state of all of the chains in the
ensemble. The idea is to consider the current state on each of a set of
parallel chains as an ensemble, and to resample using a transformation
based on optimal transport. Samplers based on optimal transport
have also been considered in \cite{el2012bayesian}.

In Section \ref{Sec:Prelim} we introduce some mathematical
preliminaries upon which we will later rely. In Section \ref{Sec:PAIS}
we present the general framework of the PAIS algorithm. In Section
\ref{Sec:adapt} we consider adaptive versions of PAIS which
automatically tune algorithmic parameters concerned with the proposal
distributions. In Section \ref{sec:AMR} we introduce the approximate
multinomial resampling (AMR) algorithm which is a less accurate but
faster alternative to resamplers which solve the optimal transport
problem exactly. In Section \ref{Sec:Num} we present some numerical
examples, before a brief conclusion and discussion in Section
\ref{Sec:Conc}.

@@

The parameter $u$ is observed
through the observation operator $\mathcal{G}:X \to\mathbb{R}^d$. 
Since observations are never perfect, we
assume that these measurements $\data$ are subject to Gaussian noise,
so that
\begin{equation}\label{eqn:obs}
	\data = \mathcal{G}(u) + \varepsilon, \qquad \varepsilon \sim \mu_{\varepsilon} = \mathcal{N}(0,\Sigma).
\end{equation}
For example, if $u$ are the rates of reactions in a chemical system, $\G$ might
return the quantities of each chemical species at a particular time, or some
summary of this information.

@@

The ensemble transform particle filter (ETPF) proposed by Reich
\cite{reich2013nonparametric} makes use of optimal transportation as
described in \cite{villani2003topics,villani2008optimal}. The transform
takes a sample of weighted particles $\{y_i\}_{i=1}^M$ from $\mu_Y$
and converts it into a sample of evenly weighted particles $\{x_i\}_{i=1}^M$
from $\mu_X$, by means of defining a coupling $T^*$ between $Y$ and
$X$. Given that a trivial coupling $T^t$ always exists in the space of
transference plans, $\Pi(\mu_{X}, \mu_{Y})$, we can find a coupling
$T^*$ which maximises the correlation between $X$ and $Y$
\cite{cotter2012ensemble}. This coupling is the solution to a linear
programming problem in $M^2$ variables with $2M-1$ constraints.
Maximising the correlation ensures that the new sample is as much like
the original sample as possible with the additional property that the sample
is evenly weighted.

@@

\subsection{Deficiencies of Metropolis-type MCMC schemes}
All MCMC methods are naively parallelisable. One can take a method
and simply implement it simultaneously over a set of processors in an ensemble. All
of the states of all of the ensemble member can be recorded, and in the time
that it takes one MCMC chain to draw $N$ samples, $M$ ensemble members
can draw $NM$ samples.

However, we argue that this is not an optimal
scenario. First of all, unless we have a lot of information about the
posterior, we will initialise the algorithm's initial state in the
tails of the distribution. The samples that are initially made as the
algorithm finds its way to the mode(s) of the distribution cannot be
considered to be samples from the target distribution, and must be thrown
away. This process is known as the burn-in. In a trivially parallelised
scenario, each ensemble member must perform this process independently,
and therefore mass parallelisation makes no inroads to cutting this cost.
